---
title: "3 Data visualisation"
output:
html_document: default
html_notebook: default
---

```{r}
suppressPackageStartupMessages(library(tidyverse))
```

# 3.2.4 Exercises

1.  Run `ggplot(data = mpg)` what do you see?
```{r}
ggplot(data = mpg)
```

There are no aesthetics or geom layers, so ggplot will not display any data.

1.  How many rows are in `mtcars`? How many columns?
```{r}
# Rows
nrow(mtcars)
# Columns
ncol(mtcars)
# Both
dim(mtcars)
```


1.  What does the `drv` variable describe?  Read the help for `?mpg` to find
out.

The type of drive that can be:

* f = front-wheel drive, 

* r = rear wheel drive, 

* 4 = 4wd

1.  Make a scatterplot of `hwy` vs `cyl`.

```{r}
ggplot(mpg, aes(x = hwy, y = cyl)) + 
  geom_point()
```


1.  What happens if you make a scatterplot of `class` vs `drv`. Why is
the plot not useful?

```{r}
ggplot(mpg, aes(x = class, y = drv)) + 
  geom_point()
```

Lots of values are overlapping and the plot does not reveal which points are overlapped and how many times.

# 3.3.1 Exercises

1.  What's gone wrong with this code? Why are the points not blue?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

The colour is specified inside the aes, while the colour "blue" does not relate to the data aesthetics. This should be:

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

1.  Which variables in `mpg` are categorical? Which variables are continuous? 
(Hint: type `?mpg` to read the documentation for the dataset). How
can you see this information when you run `mpg`?

```{r}
mpg
```

The data is R does not contain any categorical variables. This can be seen by looking at the variable types displayed under the variable names. None of them is marked as <fctr>. 

However, when reading the documentation it becomes clear that at least the following variables should be categorical: cyl, drv, fl and class.

1.  Map a continuous variable to `color`, `size`, and `shape`. How do
these aesthetics behave differently for categorical vs. continuous
variables? 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = cty))

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = cty))
```

Colour and shape will indicate how the size is spread across the displayed points on the plot. Shape can be applied only on the discreet variables, so this is not an option.

1.  What happens if you map the same variable to multiple aesthetics? 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = cty, size = cty))
```

All aesthetics will be applied at same time that often might be overkill.

1.  What does the `stroke` aesthetic do? What shapes does it work with?
(Hint: use `?geom_point`)

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), shape = 21)
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), shape = 21, stroke = 1)
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), shape = 21, stroke = 5)
```

Stroke will modify the width of the border for the shapes that have border, like for example shape 21.

1.  What happens if you map an aesthetic to something other than a variable 
name, like `aes(colour = displ < 5)`?  

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = displ < 5))
```

The aesthetics will apply on the (evaluated) expression in the same way as if it was a single variable.

# 3.5.1 Exercises

1.  What happens if you facet on a continuous variable?

```{r, eval = FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = drv, y = cyl)) +
  facet_wrap( ~ displ)
``` 

Not a good idea! R makes separate plot for each unique value of the variable.

1.  What do the empty cells in plot with `facet_grid(drv ~ cyl)` mean?
How do they relate to this plot?

```{r, eval = FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = drv, y = cyl)) +
  facet_grid(drv ~ cyl)
``` 

There are combinations that are not represented in the data set. For example, none of the observations has values like drv=f4 & cyl=f.

1.  What plots does the following code make? What does `.` do?

```{r eval = FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl)
```

The first plot is faceting drv column wise against the columns specified in the aesthetics. The second plot is faceting hwy row wise and against the columns specified in the aesthetics.

1.  Take the first faceted plot in this section:

```{r, eval = FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```

What are the advantages to using faceting instead of the colour aesthetic?

What are the disadvantages? How might the balance change if you had a larger dataset?

The data is visually separated and hence may be easier to analyse the different cases that relate to single variable value.

At the other side, this way of plotting might make it more difficult to compare cross-values.

If there is lots of data the plots could be fuller. In combination with lots of facets this can be prove difficult to read. However, by dividing lots of data in smaller groups might reveal some patterns that are difficult to see on one plot. Depending on the content there could be pro or con for faceting big data set, just like they are in a case of the small data sets.

1.  Read `?facet_wrap`. What does `nrow` do? What does `ncol` do? What other
options control the layout of the individual panels? Why doesn't
`facet_grid()` have `nrow` and `ncol` variables?

* nrow - sets number of rows on the final plot.

* ncol - sets number of columns on the final plot.

* shrink - used to shrink scales by fitting statistics of the data, rather than raw data.

* dir - direction such as horizontal and vertical

* strip.position - position labels on the chosen place

1.  When using `facet_grid()` you should usually put the variable with more
unique levels in the columns. Why?

If there is only one level, the plot will be the same as if it is not faceted

# 3.6.1 Exercises

1.  What geom would you use to draw a line chart? A boxplot? 
A histogram? An area chart?

```{r}
# Line chart
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_line()
# Boxpolt
ggplot(data = mpg, mapping = aes(x = factor(cyl), y = hwy)) +
  geom_boxplot()
# Histogram
ggplot(data = mpg, mapping = aes(x = displ)) +
  geom_histogram()
# Area chart
ggplot(data = mpg, mapping = aes(x = displ)) +
  geom_area(stat = "bin")
```


1.  Run this code in your head and predict what the output will look like.
Then, run the code in R and check your predictions.

```{r, eval = FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() + 
  geom_smooth(se = FALSE)
```

The scatter plot with displ on x and hwy on y. 

The dots are coloured by drv levels. 

The prediction lines without the confidence interval are displayed per drv level and in the same colour as the points.

1.  What does `show.legend = FALSE` do?  What happens if you remove it?  
Why do you think I used it earlier in the chapter?

```{r, eval = FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point(show.legend = FALSE)
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point()
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point(show.legend = TRUE)
```

This parameter determines whether to display the legend. It's defaulted to TRUE, so it only has to be included if the legend is not desirable. Using the legend is very useful, but it some cases is better to omit it. For example, when the plot is too busy and you need space or when you plot multiple layers and legend is repeated.

1.  What does the `se` argument to `geom_smooth()` do?

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() + 
  geom_smooth()

ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() + 
  geom_smooth(se = FALSE)
```


It draws confidence interval round the smooth line(s).

1.  Will these two graphs look different? Why/why not?

```{r, eval = FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

They will look the same because they contain exactly same data and aesthetics. The only difference is that the first version reuses code for plotting the points and smooth lines, while the second version contains explicit (but same) code base for each version.

1.  Recreate the R code necessary to generate the following graphs.

```{r echo = FALSE, fig.width = 3, out.width = "50%", fig.align = "default", message = FALSE}
require(gridExtra)

gg11 <- ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth(se = FALSE)
gg12 <- ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(aes(group = drv), se = FALSE) +
  geom_point()
gg21 <- ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() + 
  geom_smooth(se = FALSE)
gg22 <- ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(aes(color = drv)) + 
  geom_smooth(se = FALSE)
gg31 <- ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(aes(color = drv)) +
  geom_smooth(aes(linetype = drv), se = FALSE)
gg32 <- ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(size = 4, colour = "white") + 
  geom_point(aes(colour = drv))

grid.arrange(gg11, gg12, gg21, gg22, gg31, gg32, ncol=2)
```

# 3.7.1 Exercises

1.  What is the default geom associated with `stat_summary()`? How could
you rewrite the previous plot to use that geom function instead of the 
stat function?

```{r}
# Original plot
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.ymin = min,
    fun.ymax = max,
    fun.y = median
  )

# Rewritten plot
ggplot(data = diamonds) + 
  geom_pointrange(
    mapping = aes(x = cut, y = depth),
    stat = "summary",
    fun.ymin = min,
    fun.ymax = max,
    fun.y = median
  )  
```

stat_summary is statistical function that transforms data into summaries


2.  What does `geom_col()` do? How is it different to `geom_bar()`?

```{r}
ggplot(diamonds) +
  geom_bar(aes(x = cut))

ggplot(diamonds) +
  geom_bar(aes(x = cut, weight = price))

ggplot(diamonds, ) +
  geom_col(aes(x = cut, y = price))
```

geom_bar creates bar charts that shows counts (or sums of weights). It uses stat_count to get the statistics.

geom_col is used when the hights of the bar's are representing actual values. It uses stat_identity, i.e. it leaves the data as it is.

1. Most geoms and stats come in pairs that are almost always used in concert. Read through the documentation and make a list of all the pairs. What do they have in common?

They represent same statistical transformation. Here are some examples

* geom_bar & stat_count

* geom_cut & stat_count

* geom_boxplot & stat_boxplot

* geom_smooth & stat_smooth

* geom_point & stat_identity

* geom_count & stat_sum

* geom_freqpoly & stat_bin

* geom_histogram & stat_bin

* geom_density & stat_density

* geom_density_2d & stat_density_2d

* geom_bin2d & stat_bin_2d

* geom_contour & stat_contour

* geom_hex & stat_bin_hex

* geom_qq & stat_qq

* geom_quantile & stat_quantile

* geom_violin & stat_ydensity

4.  What variables does `stat_smooth()` compute? What parameters control its behaviour?

* predicted value, controled by the parameter y

* lower pointwise confidence interval around the mean, controled by the parameter ymin

* upper pointwise confidence interval around the mean, controled by the parameter ymax

* standard error, controled by the parameter se

5.  In our proportion bar chart, we need to set `group = 1`. Why? In other
words what is the problem with these two graphs?

```{r, eval = FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = ..prop..))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = color, y = ..prop..))
```

The values are grouped in equal parts. They need to be represented as proportions of the number of inspected occurrences with respect to the number of total occurrences per level.

```{r, eval = FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = ..prop.., group = 1))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = color), position ="fill")
```

# 3.8.1 Exercises

1.  What is the problem with this plot? How could you improve it?

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_point()
```

Lot of points are overlapping each other. That can be avoided by using position = "jitter" or geom_jitter

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_jitter()
```

1.  What parameters to `geom_jitter()` control the amount of jittering?

Width and hight

```{r}
# Default
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_jitter()

# Width
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_jitter(width = 0.01)
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_jitter(width = 0.5)

# Hight
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_jitter(hight = 0.1)

ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_jitter(hight = 20)

# Width and hight
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_jitter(width = 0.1, hight = 20)
```

1.  Compare and contrast `geom_jitter()` with `geom_count()`.

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_jitter()
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_count()
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_point()
```

Jitter provides insight in the real distribution by placing points with random noise so that there will be no overlapping, while count presents the density of overlapping points by increasing the size of the points. geom_point just places points on a top of each other so that it is visible where are the covered points.

1.  What's the default position adjustment for `geom_boxplot()`? Create
a visualisation of the `mpg` dataset that demonstrates it.

```{r}
# Default (dodge)
ggplot(mpg, aes(x = factor(cyl), y = hwy, colour = factor(drv))) +
  geom_boxplot()

# Other
ggplot(mpg, aes(x = factor(cyl), y = hwy, colour = factor(drv))) +
  geom_boxplot(position = "dodge")
ggplot(mpg, aes(x = factor(cyl), y = hwy, colour = factor(drv))) +
  geom_boxplot(position = "jitter")
ggplot(mpg, aes(x = factor(cyl), y = hwy, colour = factor(drv))) +
  geom_boxplot(position = "nudge")
```

Default position is dodge. Not all of the positions make sense for this type of the plot.

# 3.9.1 Exercises

1.  Turn a stacked bar chart into a pie chart using `coord_polar()`.

```{r}
ggplot(mpg, aes(x = fl, fill = drv)) +
  geom_bar(position = "stack")

ggplot(mpg, aes(x = fl, fill = drv)) +
  geom_bar() +
  coord_polar()
```

1.  What does `labs()` do? Read the documentation.

labs enables change of the default text on the plot, such as lables and titles:

```{r}
# Default
ggplot(mtcars, aes(mpg, wt, colour = cyl)) + 
   geom_point()
# Changed
ggplot(mtcars, aes(mpg, wt, colour = cyl)) + 
   geom_point() + 
  labs(colour = "Cylinders") + 
  labs(x = "Mileage per gallon", y = "Weight") + 
  labs(caption = "(based on data from mpg data set)") +
  labs(title = "Millage per gallon vs. weight", subtitle = "Scatter plot")
```

1.  What's the difference between `coord_quickmap()` and `coord_map()`?

```{r}
library(maps)
library(mapproj)
nz <- map_data("nz")
# Prepare a map of NZ
nzmap <- ggplot(nz, aes(x = long, y = lat, group = group)) +
  geom_polygon(fill = "white", colour = "black")
# Plot it in cartesian coordinates
nzmap + labs(caption = "Default")

# With correct mercator projection
nzmap + coord_map() + labs(caption = "coord_map")
# With the aspect ratio approximation
nzmap + coord_quickmap() + labs(caption = "coord_quickmap")
```

coord_map wil convert the map to display area in realistic proportions. 

coord_quickmap is using less calculations for the conversion so it is quicker to display but less precise.


4.  What does the plot below tell you about the relationship between city and highway mpg? Why is `coord_fixed()`, important? What does `geom_abline()` do?

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() + 
  geom_abline() +
  coord_fixed()
```

coord_fixed does not resize the x or y to fit the data on the available space in the best manner. It keeps the ratio 1:1

geom_abline draws a line. If no parameters are provided, the line will be specified as x = y but it can be changed by specifying slope (angle) and intercept (y point when x = 0)

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() + 
  geom_abline() +
  coord_fixed(ratio = 1/5)
```
