---
title: "4 Workflow: basics"
output:
  html_notebook: default
---

```{r setup, echo = FALSE, message = FALSE}
library(tidyverse)

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE
)
```

# 4.4 Practice

1.  Why does this code not work?

```{r 4.4.1, eval = FALSE}
my_variable <- 10
my_var覺able
#> Error in eval(expr, envir, enclos): object 'my_var覺able' not found
```

Look carefully! (This may seem like an exercise in pointlessness, but training your brain to notice even the tiniest difference will pay off when programming.)

**

The letter i in first statement does not match the character 覺 that is used in the second line, so R will not be able to find my_var覺able.

***

2.  Tweak each of the following R commands so that they run correctly:


```{r 4.4.2, eval = FALSE}
library(tidyverse)

ggplot(dota = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

fliter(mpg, cyl = 8)
filter(diamond, carat > 3)
```

**

```{r}
# replace dota by data
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

# replace = by ==
filter(mpg, cyl == 8)

# diamond misses s at the end
filter(diamonds, carat > 3)
```

***

3.  Press Alt + Shift + K. What happens? How can you get to the same place using the menus?

**

The list with shortcut's popes up. Menu/Help/Keyboard Shortcuts Help does the same.
